---
title: "Parasite Transmission by Adult Mosquitoes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parasite Transmission by Adult Mosquitoes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(ramp.xds)
library(viridisLite)
```

```{r, echo=F}
#devtools::load_all()
```

# $\oplus$  

The basic mosquito models in **`ramp.xds`** are variants of George Macdonald mathematical models, developed with help from P. Armitage. The variables for all the models in **`ramp.xds`** are mosquito densities, and the default model is from Aron & May, 1982.

## George Macdonald 

George Macdonald's papers published from 1952 transformed the quantitative study of medical entomology and malaria transmission by mosquitoes: 

+ 1952 - [*The analysis of the sporozoite rate*](https://pubmed.ncbi.nlm.nih.gov/14958825/) presented a formula for the sporozoite rate. Macdonald's analysis demonstrated the importance of mosquito survival.  

+ 1952 [*The analysis of equilibrium in malaria*](https://pubmed.ncbi.nlm.nih.gov/12995455/) introduced the *basic reproduction rate* of malaria with a compact set of parameters. 

Macdon
In Macdonald's notation (and some of his words): 

+ $m=$ the ratio of anopheline vectors to humans 

+ $a=$ the average number of humans bitten by one mosquito in one day. 

+ $b=$ the proportion bites by infectious mosquitoes that cause an infection.  

+ $p=$ the probability a mosquito surviving through one whole day.  

+ $n=$ the time taken for completion of the extrinsic cycle

The mathematical models that are now commonly called the *Ross-Macdonald* model typically combine Ronald Ross's model for human infections, an *SIS* compartmental model, with an *SEI* model for mosquito infections. As a differential equation, Macdonald's model was almost certainly a delay differential equation. 

**`ramp.xds`** has adopted as its default model (`MYZname = "RM"`) a set of equations published by Joan Aron and Bob May in 1982. In the following, we discuss models for mosquito ecology and infection dynamics in **`ramp.xds`** 

## `RM`

Let $M$ denote the density of adult mosquitoes, $\Lambda(t)$ the emergence rate of adult mosquitoes from aquatic habitats, and $g$ the adult mosquito mortality rate. Mosquito population density is described by an equation: 

$$\frac{dM}{dt} = \Lambda(t) - g M$$
If $\Lambda$ is constant, then the equation has a steady state: 

$$\bar M = \Lambda/g$$ 

The dynamics of infection depend on a quantity that we will call $\kappa(t)$ denoting the net infectiousness (NI) of humans, or the probability a mosquito becomes infected after blood feeding on a human. Let $f$ denote the overall blood feeding rate, and let $q$ denote the fraction of blood meals on humans (*a.k.a* the human fraction). Let $Y$ denote the density of mosquitoes that are infected with parasites, and its dynamics are described by: 

$$\frac{dY}{dt} = fq\kappa(M-Y) - gY$$
Let $Z$ denote the density of mosquitoes that have sporozoites in their salivary glands, and let $\tau$ denote the time elapsed between the blood meal that infects a mosquito and the point in time when a surviving mosquito has sporozoites in its salivary glands.  The probability a mosquito has survived $\tau$ days is $e^{-g\tau}$, and we let a subscripted $\tau$ denote the value of a variable at time $t-\tau.$ The density of infectious mosquitoes is modeled by:  

$$\frac{dZ}{dt} = e^{-g \tau} fq\kappa_\tau(M_\tau-Y_\tau) - gZ$$
If $M$ and $\kappa$ have reached the steady state, then the fraction of infected mosquitoes, $y=Y/M$ is given by:

$$y = \frac{fq\kappa}{g+fq\kappa}$$ 

and the fraction of *infectious* mosquitoes (*i.e.* with sporozoites in their salivary glands), is called the sporozoite rate,

$$z = e^{-g\tau} \frac{fq\kappa}{g+fq\kappa}$$ 
This is equivalent to Macdonald's formula. 

In **`ramp.xds`** this set of equations is implemented. This sets up and solves a model with $\kappa$ and $\Lambda$ passed as constant values, and a human population density of $H=10.$ 

```{r}
rm = xds_setup(Xname = "trivial", Xopts = list(kappa=0.1), Lname = "trivial", Lopts = list(Lambda=5), HPop=10)
rm = xde_solve(rm, Tmax=100)
```

```{r, fig.width=5, fig.height=3.5}
xds_plot_M(rm)
xds_plot_YZ(rm, add_axes=FALSE)
```




### Macdonald's Formula

If we wanted to replicate Macdonald's equations, we would let $a=fq$, $p = e^{-g}$ or equivalently $g = -\ln p$, $\kappa = x,$ and $n=\tau$ so: 

$$e^{-g\tau} \frac{fq\kappa}{g+fq\kappa} = p^n \frac{ax}{ax - \log_{e}p}$$

### Spatial Dynamics 

**`ramp.xds`** can also model mosquito spatial dynamics, as described in [Spatial Dynamics of Malaria Transmission (2023) *PLoS Computational Biology*](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1010684). In the spatial formulations, a landscape is sub-divided into a set of $n_p$ patches. The variables $M$ and $Y$ and $Z$ are vectors of length $n_p,$ and the paramger $g$ is replaced by a mosquito demographic matrix $\Omega$ that describes dispersal. The matrix $\Omega$ combines several parameters:

+ $g$ - a vector describing patch-specific dispersal

+ $\sigma$ - a vector describing patch-specific emigration rates 

+ $\mu$ - a vector describing the fraction of mosquitoes *lost* through migration, including mortality and dispersal out of the system

+ $\cal K$ - a dispersal matrix

$$\Omega = \mbox{diag}\left(g\right) - \left[\mbox{diag}\left(1-\mu\right)-{\cal K} \right] \cdot
\mbox{diag}\left(\sigma\right)$$

## Variants 

**`ramp.xds`** also includes two variants of Macdonald's model:  

+ `si` is a simpler version of the model above 

+ `sei` is a compartmental model 


### `si`

A simple variant of Macdonald's model ignores the delay for the EIP. The equation for $M$ and $Y$ are as above, but 

$$Z = e^{-g\tau} Y$$ 
```{r}
si = xds_setup(MYZname = "si", Xname = "trivial", Xopts = list(kappa=0.1), Lname = "trivial", Lopts = list(Lambda=5), HPop=10)
si = xde_solve(si, Tmax=100)
```

```{r, fig.width=5, fig.height=3.5}
xds_plot_YZ_fracs(si)
xds_plot_YZ_fracs(rm, add_axes=FALSE, llty=2)
```


### `sei`

In a model that is equivalent to the standard SEI compartmental model (under a change of variables), the equations for $M$ and $Y$ are as above, but the fraction infectious is: 

$$\frac{dZ}{dt} =  (Y-Z) / \tau - g Z$$  

```{r}
sei = xds_setup(MYZname = "sei", Xname = "trivial", Xopts = list(kappa=0.1), Lname = "trivial", Lopts = list(Lambda=5), HPop=10)
sei = xde_solve(sei, Tmax=100)
```

```{r, fig.width=5, fig.height=3.5}
xds_plot_YZ_fracs(rm, llty=2)
xds_plot_YZ_fracs(sei, add_axes=FALSE)
```

## Comparing Models

```{r, fig.width=5, fig.height=3.5}
clrs = turbo(4)[-2]
xds_plot_Z_fracs(rm, Zclrs=clrs[1])
text(80, 0.15, "RM", col = clrs[1])
text(80, 0.11, "si", col = clrs[2])
text(80, 0.07, "sei", col = clrs[3])
xds_plot_Z_fracs(si, add_axes=FALSE, Zclrs = clrs[2])
xds_plot_Z_fracs(sei, add_axes=FALSE, Zclrs = clrs[3])
```

```{r}
rm1 <- rm 
rm1$Lpar[[1]]$season = function(t){1+sin(2*pi*t/365)}
rm1$Lpar[[1]]$trend = function(t){exp(-t/1000)}
rm1 <- xde_solve(rm1, Tmax=1500)
xds_plot_M(rm1)
```


## Exogenous Forcing


