---
title: "Basic Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Previous: [*Getting Started.*](GettingStarted.html) Next: [*Working with Models*.](Working.html) 
Also, see [**SimBA**](https://faculty.washington.edu/smitdave/simba/index.html)

*** 


**`ramp.xds`** is based on a mathematical framework for model building that is modular, flexible and extensible. Using the built-in functions, it's comparatively easy to build highly *realistic* models by starting simple and adding features one at a time. We call this *progressive model building.* 

In the vignette [Getting Started](GettingStarted.html), we introduced the function `xds_setup().` From a software design perspective, a setup interface can't possibly anticipate the needs of every end user without constraining possibilities. The point of `xds_setup()` is to help build models up to a certain level of complexity. Instead, we have adopted a system for *progressive model building* based on a mathematical framework and the idea of *scaling complexity.* 

Progressive model building is a practical method for building complex models in two phases:  

+ **Basic Setup** is handled by the function `xds_setup().` It's designed to help a user get started with a properly formulated interface that can handle spatial dynamics, structured aquatic habitats, and a stratified human population.

+ **Advanced Setup** handles advanced features, including exogenous forcing, vector control, and other modes of malaria control, human demography and aging, and host population stratification. Advanced setup is described in [**SimBA**](https://faculty.washington.edu/smitdave/simba/index.html)

In the following, we introduce **Basic Setup**

## Modularity and Etensibility 

**`ramp.xds`** was developed around a mathematically rigorous framework that takes advantage of the natural modularity of mosquito-transmitted pathogens. The core functions handle [**spatial dynamics**](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1010684), and the software was designed to model malaria as a *changing baseline modified by control.* The framework is structurally flexible and extensible.  


**Basic setup** creates a model that sets up the basic structural elements for the spatial dynamics, and that sets up the dynamical modules. A model returned by basic setup is designed to be modified with the potential to develop models of unbounded realism and complexity. To make this work, the software built in ports and junctions that could handle a large set of advanced features, but most of these features would be turned off for basic setup. 

### Modularity

The framework is designed around five dynamical components organized into three chunks: 

1. Human / vertebrate host ecology, infection and immunity, called the **XH**-component 

    + human (or host) demography (**H**-component) including births, deaths, aging and migration;

    + malaria epidemiology (**X**-component), in the narrow sense; or more generally, the dynamics of infection and immunity in the host population;  

2. Adult mosquito ecology and infection dynamics, called the **MYZ**-component 

    + mosquito ecology: the **M**-component 

    + mosquito infection dynamics: the **Y**-component 
    
    + parasite development: the **Z**-component 

3. Aquatic mosquito ecology, called the **L**-component 

These three chunks are connected by two interfaces. The first two chunks are connected through the rigid interface describing blood feeding and transmission. The last two chunks are connected through egg laying and emergence. 

In general, A *module* describes an element in the model library that executes all the functions in one of the core chunks. Since the parameter values can be changed, each *module* defines a *model family* in the form of a system of differential or difference equations. A model family is defined by a set of states and state transitions (*i.e.* by the topology of the graph) for a single population (`nStrata=1`). 

For example, the `SIS` module for the **X**-component is the *SIS* compartmental model family for human infections. It's called a model family because each unique set of parameters would define a different *model.* 

In its simplest form, each model family represents the state of a single population, but any model can replicate those states and sub-divide a single population into sub-populations with different parameters. These are population strata for the **X**-component; adult mosquitoes are found in patches for the **MYZ**-component; immature mosquitoes in aquatic habitats for **L**-component. In **`ramp.xds`** it is comparatively simple to configure a model with multiple strata, patches, and habitats. 

## Basic Setup  

*For our purposes, **Basic Setup** encompasses the configurable options in `xds_setup().`* Each model returned by `xds_setup` is characterized by:

+ A *model family* for each dynamical component, including model parameters. The model object also stores initial values. 

+ The *model structure,* including the number of patches, the number of habitats and their membership in the patches, and the number of human population strata and their residency within the strata. 

+ Functions that define exogenous variables and functional responses that drive exogenous forcing by weather and malaria control.  

The software supports a modular framework for model building that is highly flexible and extensible. It's capable of building very realistic models, but progressive setup embraces a model-building philosophy of starting simple and adding realism or complexity one step at a time. 

The demo in [Getting Started](GettingStarted.html) sets up one version of the Ross-Macdonald model. It illustrates how the software has streamlined the process of building fairly simple models. The following sections describe how to make **`ramp.xds`** build models that are much more complicated and realistic. 

As described in [Getting Started](GettingStarted.html), the `xds_setup()` function **`ramp.xds`** has assigned default options for everything. Each model must configure three dynamical components: human epidemiology; adult mosquito ecology and infection dynamics; and aquatic mosquito ecology. Each one of these three dynamical component is configured by specifying a model name (as a string) and options, passed as a named list, to override the defaults for the model: 

+ `Xname` dispatches a module for the **X** component. The setup default model is `Xname="SIS"` and default parameters for the `SIS` model are defined in `create_Xpar_SIS.` To replace the default values at setup, a named list can be passed as `Xopts = list(...)` and the named values in `Xopts` will be 
he values of named parameters and the initial values of named variables instead of the defaults. (The human demographic module for the **H**-component is set up as an advanced option.) 

+ `MYZname` dispatches a module for the **MYZ** component. The setup default model is `MYZname="macdonald"` and default parameters for the `macdonald` model are defined in `make_MYZpar_macdonald.` To replace the default values at setup, a named list can be passed as `MYZopts = list(...)` and the named values in `MYZopts` will set the values of named parameters and the initial values of named variables instead of the defaults.  

+ `Lname` dispatches a module for the **L** component. The setup default model is the trivial model `Lname = "trivial"`. It has no variables, but it passes the outputs -- emerging adults -- as a parameter. To replace the default values at setup, a named list can be passed as `Lopts = list(...)` and the values in `Lopts` will set the values of named parameters and the initial values of named variables instead of the defaults.  

### Defaults

```{r}
library(ramp.xds)
```

```{r, echo=F}
#devtools::load_all()
```


Thu function call: 

```{r, eval=F}
model1 = xds_setup()
```

returns the same values as the function call: 

```{r, eval=F}
model2 = xds_setup(Xname = "SIS", 
                   MYZname = "macdonald", 
                   Lname = "trivial")
```

### Structural Parameters 

Structural parameters describe things like the number of patches, the number of population strata, the number of aquatic habitats, the number of host species, and the number of vector species. `xds_setup` sets up models with one host species and one vector species, users can configure a malaria landscape with *spatial dynamics,* structured *aquatic habitats,* and *multiple human / host* population strata. Basic setup was developed to ensure that all interfaces are set up properly when there are multiple patches, multiple habitats, multiple population strata, and malaria importation. The software and documentation follow the mathematical framework formulas and (with small changes) the notation in [Wu SL, *et. al.* 2023](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1010684). The following structural parameters can be set through `xds_setup:`
    
- **`nPatches`** or $n_p$ is the number of patches in the model. 
    
- **`membership`** is set up by passing a vector with the identity of patch where each habitat can be found, and it sets up the habitat membership matrix, $Q$. Setup sets the value of a variable describing the number of habitats, $n_q$ = **`nHabitats <- length(membership)`.** One configurable option is **`searchQ`** with the habitat search weights. 
    
- **`HPop`** is a vector with human population density, and **`residence`** is a vector describing the location of the patch where each stratum resides.  Setup checks that `length(HPop) == length(residence)` and sets the number of strata, $n_h$ or **`nStrata <- length(residence)`.** It is also possible to configure heterogeneous biting at the command line by passing **`searchH`** 

### Spatial Dynamics

- If `nPatches >1` then other setup-configurable options are to pass:
    
    - a matrix $K$ or `calK` describing mosquito dispersal among the patches.  
          
    - a matrix $\Theta$ or `TimeSpent` can be passed to configure human time spent  

To learn more, we suggest taking a look at the [5-3-4 Vignette](https://faculty.washington.edu/smitdave/simba/model534.html), a model with 5 aquatic habitats, 3 patches, and 4 human population strata.  

