---
title: "Bed Nets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bed Nets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(ramp.xds)
```


```{r}
mod <- xds_setup(MYZname = "SI")
```

```{r}
cov_opts <- list(
  mean = 0.5, 
  F_season = function(t)
    {ifelse(t < 0, 0, (sin(2*pi*(t-365/4) / 365) + 1))}
)
```

```{r}
tt = seq(0:730)
with(cov_opts, plot(tt, mean*F_season(tt), type = "l"))
```


```{r}
mod <- xds_setup_bednets(mod,
     coverage_name = "func", coverage_opts = cov_opts, 
     effectsizes_name = "lemenach")

```

```{r}
tt = seq(0:730)
with(mod$bednets$coverage, plot(tt, mean*F_season(tt), type = "l"))
```
Every model for  

- `VectorControl`
    - `BedNet`
        -  `DistributeBedNets`
        -  `OwnBedNets`
        -  `UseBedNets`
        -  `BedNetEffects`
- `VectorControlEffectSizes`
    - `BedNetCoverage`
    - `BedNetEffectSizes`

setup_bednet = function(pars, 
    dist_name = `basic`,
    distribute = list(),
    own_name = `basic`,
    own = list(), 
    use_name = `basic`,
    use = list(), 
    effects_name = `basic`,
    effects = list(), 
    coverage_name = `basic`,
    coverage = list(),
    es_name = `basic`,
    effect_sizes =list()
)
