---
title: "Mosquito Bionomics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mosquito Bionomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In **`ramp.xds,`** mosquito bionomics are handled as a junction: each parameter describing adult mosquito ecology or infection dynamics is handled as a *port:*

+ During *basic setup,* in **`ramp.xds`**, the parameters are assigned constant values. 

+ Some parameters can be time-dependent to model forcing by weather or resource availability. Forcing is an *advanced setup* option. Advanced setup options are in **`ramp.forcing`**

+ Various modes of vector control are handled as another set of *advanced setup* options: advanced setup options for vector control are in **`ramp.control.`**  Vector control effects can be handled in one of two ways: 

    - vector control can affect the availability of resources, and the effects are computed through functional forms in response to available resources
    
    - vector control effect sizes can be computed.  

Each module in **`ramp.xds`** or in the satellite package **`ramp.library`** must be capable of handling time-varying parameters, or it should report that does not. For example, 
the module `macdonald` is not capable of handling either forcing or vector control. We would say that it is a strictly autonomous model: those capabilities are not in its *skill set.* To handle forcing and vector control, we developed `GeM.`

## Mosquito Dispersal

In **`ramp.xds`**, mosquito dispersal is understood conceptually as the result of a single flight bout. Regardless of how well a mosquito can fly without the help of wind, wind ends up being an important factor affecting mosquito dispersal, and it is possible for a mosquito to travel reasonably long distances with the help of the wind. 

In adult mosquito modules, mosquito dispersal is described by 
a square matrix \eqn{K} with `nPatches` \eqn{(=n)} rows and columns. By convention,
all \eqn{K} matrices have the form:
$$
  K = \left[ \begin{array}{ccccc}
  -1 & k_{1,2} & k_{1,3} & \cdots & k_{1, n} \\
  k_{2,1} & -1 & k_{2,3} & \cdots & k_{2, n} \\
  k_{3,1} & k_{3,2} & -1 & \cdots & k_{3, n} \\
  \vdots & \vdots & \vdots & \ddots  & \vdots \\
  k_{n,1} & k_{n,2} & k_{n,3} & \cdots & -1 \\
  \end{array} \right]
$$
The values of the elements \eqn{k_{i,j}} are thus 
constrained such that there is no net dispersal loss from the system: \eqn{\forall i,}
$$\sum_j k_{i,j} = 1.$$ 
The columns of \eqn{K} sum up to zero. A dispersal matrix, $K,$ thus describes the destinations of 
emigrating mosquitoes that survive and stay in the system. Mosquito loss from a system due to emigration is handled elsewhere. 

## Mosquito Demography

Mosquito survival and dispersal is described by a *demographic matrix,* denoted \eqn{\Omega.} It is 
computed using several parameters:

+ $g$ --- the daily mosquito mortality rate
+ $\sigma$ --- the patch emigration rate
+ $\mu$ --- the rate of emigration-related loss
+ $K$ --- a dispersal matrix (see above)

These terms were designed to make it possible to develop and calibrate models for mosquito mortality that make it easy to specify *how* mosquitoes are lost from a system. The term $\mu$ describes loss from a system that is associated with emigration: The emigration rate is $\sigma,$ but only a fraction $\sigma(1-\mu)$ land in an other patch. The remaining $\sigma \mu$ are lost from the system. Mosquito mortality and migration are computed as a single object, $\Omega,$ where
 
$$
 \Omega = \mbox{diag} \left( g + \sigma \mu \right) - K \cdot \mbox{diag} \left( \sigma \left(1-\mu\right) \right)
$$
and in general: 

$$ \frac{dM}{dt} = \Lambda - \Omega \cdot M$$ 


In delay differential equations with a constant EIP ($\tau$), 
survival and dispersal through the EIP is given by: 
$$
\Upsilon = e^{-\Omega \tau} 
$$
which can be computed in R using `expm.`

## Setup 

See `mosquito_dispersal.R` 

```{r}
library(ramp.xds)
```

By default, $\sigma=\mu=0$ and $K$ is a matrix of all zeros.

```{r}
mod <- xds_setup(nPatches=3)
get_K_matrix(mod)
```

### `herethere`

Another simple option is `herethere` that distributes mosquitoes to every other patch with equal probability. The $K$ matrix can be reconfigured after setup using `setup_K_matrix`

```{r}
mod <- setup_K_matrix("herethere", mod) 
get_K_matrix(mod)
```
The same options can be passed as a named list `Koptions` during basic setup: 

```{r}
mod <- xds_setup(nPatches=6, Koptions = "herethere")
get_K_matrix(mod)
```
### `xy`

If the patches have locations, given by a pair of vectors $x$ and $y,$ each one of length `nPatches` and passed concatenated as `xy`, then a dispersal matrix $K$ is computed by passing a *kernel* function that weights by distance. Setup calls `make_K_matrix`

```{r}
x = y = c(1:5)
xy = cbind(x=x,y=y)
F_d = function(d){exp(-d^1.8)}
K <- make_K_matrix_xy(xy, F_d)
round(K*1000)/1000
```
```{r}
mod <- xds_setup(nPatches=5, Koptions = list(name = "xy", xy=xy, ker=F_d))
get_K_matrix(mod)
```
### `as_matrix`

Finally, if the user wants to form some other kind of matrix, it can be passed directly as `Koptions`: 

```{r}
mod <- xds_setup(nPatches=5, Koptions=K)
get_K_matrix(mod)
```

### `no_setup` 

Internally, the default $K$ matrix is set up with the option `no_setup,` which sets the $K$ matrix to all zeros. 

```{r}
mod <- setup_K_matrix("no_setup", mod)
get_K_matrix(mod)
```



