---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Next: [Basic Setup](BasicSetup.html). 
Also, see [**SimBA**](https://faculty.washington.edu/smitdave/simba/_site/index.html)

*** 

# 

**`ramp.xds`** was developed to reduce the costs of building, solving, analyzing, and using models describing the epidemiology, transmission dynamics and control of malaria and other mosquito-transmitted pathogens. This vignette gives a demonstration and a basic introduction. 


## Installation 

The software was developed in RStudio. To download the R package, run these commands: 

```{r, eval=F}
library(devtools)
devtools::install_github("dd-harp/ramp.xds")
```

Then load **`ramp.xds`**: 

```{r}
library(ramp.xds)
```

```{r, echo=F}
#devtools::load_all()
```

Additional modules are available in the companion R-package [**`ramp.library`**](https://dd-harp.github.io/ramp.library/)

## Demo

**`ramp.xds`** makes it easy to build and use models for the transmission dynamics and control of malaria or other mosquito-borne pathogens. 
Model building in **`ramp.xds`** starts 
with a function called `xds_setup().`
All arguments in `xds_setup()` are assigned default values, so calling `xds_setup()` with no arguments returns a fully configured model with all the default settings. 

```{r}
model <- xds_setup()
```

The object `model` returned by `xds_setup()` is called a **xds model** object: it is a fully defined model that can be solved by `xds_solve()`. 

### Defaults 

The arguments are stored and can be easily retrieved. The name of the module for human / host infection and immune dynamics is stored as `Xname.` The default module is the `SIS` compartmental model:

```{r}
model$Xname
```

`SIS` was set up with default parameter values, that can be viewed with the function `get_Xpars`: 

```{r}
get_Xpars(model)
```
Functions like `get_Xpars` are designed to help users get to know the model oject. In this case, the parameters are stored as `model$Xpar[[i]],` for the $i^{th}$ species. Knowing the names, the parameters can then be viewed (or changed):

```{r}
model$Xpar[[1]]$r
```

Similarly, initial values were assigned default values at setup:

```{r}
get_Xinits(model)
```

The default parameters and initial values can be over-written at setup by passing `Xopts = list(...)` 
with different values: 

```{r}
Xo = list(b=1/150, c=0.2, b=0.6, S=3, I=2)
model1 <- xds_setup(Xopts = Xo)
get_Xpars(model1)
```

One thing about overwriting the defaults is that the variable $S$ in this model is computed in relation to the total population size $H$, which plays an important role in setting up the *Blood Feeding* interface, so the $S=3$ above was ignored.

```{r}
get_Xinits(model1)
```

To change the human population density, use the `set_H` function. 

```{r}
model1 <- set_H(1004, model1)
get_Xinits(model1)
```

The name of the default module for adult mosquito ecology and infection dynamics is called `MYZname.` The default module is Macdonald's model: 

```{r}
model$MYZname
```

The design pattern is followed: once again, parameters can be viewed with `get_MYZpars` and the method for setting parameter values by passing a named list of values to `MYZopts` in `xds_setup()` 

The name of the default module for aquatic mosquito ecology is called `Lname.` The default module is the `trivial` module: 

```{r}
model$Lname
```

For the `trivial` module, arguments passed at the command line configure functions that pass values to other dynamical components. 

The design pattern is followed: parameters can be viewed with `get_Lpars` and the method for setting parameter values by passing a named list of values to `Lopts` in `xds_setup()` 

### Solving 

After setting up, `model` has an empty placeholder for outputs:

```{r}
model$outputs 
```

Calling `xds_solve(model)` returns solutions to the *initial value problem(s):* `xds_solve` calls `deSolve,` and the values of the dependent variables at a set of time points are attached as `model$outputs.` 

There are three ways to pass the argument `time` to `deSolve`: first, `xds_solve` can set up a vector of time points `seq(0, Tmax, by = dt)` that are configurable with the arguments `Tmax` and `dt;` second, if no arguments are passed for time, then it uses the defaults `Tmax=365` and `dt=1;` third, `deSolve` will use the optional argument `times` if it is supplied. All four of the following do the same thing: 

```{r, eval=F}
model <- xds_solve(model, times = 0:365) 
model <- xds_solve(model, times = 0:365, Tmax=730, dt=5) 
model <- xds_solve(model, Tmax=365, dt=1) 
```

```{r}
model <- xds_solve(model) 
```

Note `xds_solve` requires an **`xds`** *model object* and that it returns an **`xds`** *model object.* In this case, the function passes `model` and the return value replaces `model.` Before `xds_solve,` `model$outputs` was an empty list. After solving, the outputs are stored as `model$outputs`: 

```{r}
names(model$outputs)
```

The outputs from solving a system of equations include:

+ `time` -- are the time points when the values of the dependent variables were evaluated
+ `last_y` -- is the dependent variables at the last time point, the last state of the system  
+ `orbits` -- are the solutions with the values of the dependent variables parsed by name, accessible separately for each dynamical component 
+ `terms` -- dynamical terms describing transmission, including the EIR 
+ `deout` -- the raw (unparsed) outputs  

Note that the `time` holds the default values of the independent variable time where we solved the initial value problem for the dependent variables. 

```{r}
head(model$outputs$time, 5)
tail(model$outputs$time, 3)
```

When `xds_solve` is called, the **`xds` model object** it returns has replaced any old values. If nothing has changed, then the outputs will be identical. 

On the other hand, if anything has changed, any old outputs will get replaced. If the outputs need to be saved for future analyses, the user will probably have to write a wrapper function that extracts and saves it.

### Parsing 

To make it easy to deal with the outputs, the dependent variables are parsed and returned in named lists. `xde_solve()` also computes some standard *terms* that are likely to be of interest.  

Functions like `get_EIR()` make it easy to examine or save a set of standard outputs without delving into the details. 

```{r}
head(get_EIR(model), 3)
```

### Visualizing

**`ramp.xds`** includes functions to plot standard outputs:

- `xds_plot_X` plots the density of infected individuals

- `xds_plot_M` plots the density of adult female mosquitoes in each patch

- `xds_plot_Y` plots the density of infected adult female mosquitoes in each patch

- `xds_plot_Z` plots the density of infectious adult female mosquitoes in each patch

- `xds_plot_PR` plots the true prevalence of infection in the human / host population 

- `xds_plot_EIR` plots the EIR for each one of the human / host population strata

```{r, fig.height=5, fig.width=6, fig.cap = "**Figure 1:** Plotting standard outputs" }
par(mfrow = c(2,2))
xds_plot_X(model)
xds_plot_M(model)
xds_plot_Y(model, add=T)
xds_plot_Z(model, add=T)
xds_plot_PR(model)
xds_plot_EIR(model)
```

### Analysis

**`ramp.xds`** has built-in functions that compute some quantities of interest, including stable states or stable orbits. 


## Trivial Models and Trace Functions

The software was designed to be fully modular, so that any module for one dynamical component can be replaced with another. The models included in **`ramp.xds`** are just enough to illustrate its the features ensure that the software is working as it should.

A basic requirement for full modularity is that every dynamical component must include a *trivial* module. In the `trivial` module, no variables are defined, but the dynamical terms needed by related dynamical components are passed by a configurable *trace* function with a generic form:

```{r, eval=F}
mean*F_season(t)*F_trend(t)
```

The default for the $\cal L$ dynamical component is the `trivial` module, where the `mean` is is called `Lambda.`

It is easy to change the `trace` function. First, we define `F_season` and change the mean value of `Lambda.` 
The generic format for seasonal functions is `F_season(t, phase, season_opts),` where `phase` sets the phase (time of year), and `season_opts` is a set of configurable parameter values:

```{r}
 F_sp = function(t){(1+sin(2*pi*t/365))} 
```

Having defined `F_sp,` we now configure a list of options:

```{r}
Lo = list(
       Lambda=500, 
       F_season=F_sp
      )
```

We use the first model as a template for the new model, but we assign the return value a new name `model_1` so that the original one still exists: 

```{r}
model_1 <- setup_Lpar("trivial", model, 1, Lopts=Lo)
```

We want to change the initial values of `model_1.` Since `model_1` was copied from `model,` the last values from solving it are available, and we can use a function `last_to_inits` to set the initial values.

```{r}
model_1 <- last_to_inits(model_1)
```

Now, we solve it over a three year period

```{r}
model_1 <- xds_solve(model_1, Tmax=365*3, dt=5) 
```

and plot the results. 

```{r, fig.height=2.6, fig.width=6, fig.cap = "**Figure 2:** Outputs with Seasonal Forced Emergence using a Trace Function" }
par(mfrow = c(1,2))
xds_plot_X(model_1)
xds_plot_M(model_1)
xds_plot_Y(model_1, add=T)
xds_plot_Z(model_1, add=T)
```


A useful place to keep reading is [Progressive Model Building](ModelBuilding.html)

For more information about trace functions, see the vignette [Trace Functions](https://dd-harp.github.io/ramp.xds/articles/TraceFunctions.html) and [Progressive Model]
