---
title: "Getting Started with `ramp.xds`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with `ramp.xds`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 

This vignette is a very brief introduction to using **`ramp.xds`**

## Installation 

```{r, eval=F}
library(devtools)
devtools::install_github("dd-harp/ramp.xds")
```

```{r}
library(ramp.xds)
```

```{r}
#devtools::load_all()
```

## Demo

The software is designed to make it easy to set up models. The function `xds_setup()` returns a fully configured model: 

```{r}
model <- xds_setup()
```

```{r}
model <- xde_solve(model) 
```

```{r}
xds_plot_PR(model)
```
```{r}
xds_plot_EIR(model)
```

## Setup  

**`ramp.xds`** includes setup functions, called `xds_setup` with default options. Each dynamical component is configured by specifying a model name (as a string) and options, passed as a named list, to override the defaults for the model: 

+ `Xname` dispatches a model family for human parasitic infections and immunity, $\cal X.$ The setup default model is `Xname="SIS"` with default parameters defined in `create_Xpar_SIS.`

+ `MYZname` dispatches a model family for adult mosquito ecology and parasite infection dynamics, $\cal MYZ.$ The setup default model is `MYZname="RM"` with default parameters defined in `create_MYZpar_RM.` 

+ `Lname` specifies a model family for aquatic mosquito ecology, $\cal L.$  The setup default model is the trivial model `Lname = "trivial"`.




