---
title: "Hybrid MoI (Multiplicity of Infection) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrid MoI (Multiplicity of Infection) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is a hybrid model which tracks the mean multiplicity of infection (superinfection) in two compartments. The first, $m1$ is all infections, and the second $m2$ are apparent (patent) infections. Therefore $m2$ is "nested" within $m1$. It is a "hybrid" model in the sense of [Nåsell (1985)](https://link.springer.com/book/10.1007/978-3-662-01609-1).

# Differential Equations

The equations are as follows:

$$
\dot{m_{1}} = h - r_{1}m_{1}
$$
$$
\dot{m_{2}} = h - r_{2}m_{2}
$$
Where $h = b EIR$, is the force of infection. Prevalence can be calculated from these MoI values by:

$$
x_{1} = 1-e^{-m_{1}}
$$
$$
x_{2} = 1-e^{-m_{2}}
$$
The net infectious probability to mosquitoes is therefore given by:

$$
x = c_{2}x_{2} + c_{1}(x_{1}-x_{2})
$$

Where $c_{1}$ is the infectiousness of inapparent infections, and $c_{2}$ is the infectiousness of patent infections.

# Equilibrium solutions

One way to proceed is assume that $m_{2}$ is known, as it models the MoI of patent (observable) infections. Then we have:

$$
h = r_{2}/m_{2}
$$
$$
m_{1} = h/r_{1}
$$
We can use this to calculate the net infectious probability, and then $\kappa = \beta^{\top} \cdot x$, allowing the equilibrium solutions of this model to feed into the other components.

# Example

```{r, message=FALSE, warning=FALSE}
library(ramp.xds)
library(deSolve)
library(data.table)
library(ggplot2)
```

```{r, echo=FALSE} 
#devtools::load_all()
```

Here we run a simple example with 3 population strata at equilibrium. We use `ramp.xds::make_parameters_X_hMoI_xde` to
set up parameters. Please note that this only runs the human population component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

We use the null (constant) model of human demography ($H$ constant for all time). While $H$ does not appear in the equations above, it must be specified as the model of bloodfeeding ($\beta$) relies on $H$ to compute consistent values.

# The Long Way

Here, we build a model step-by-step. 

```{r}
nStrata <- 3
H <- c(100, 500, 250)
residence = rep(1,3) 
params <- make_xds_object("xde", "cohort", "ode", 1, 1, residence, H)
```


```{r}
b <- 0.55
c1 <- 0.05
c2 <- 0.25
r1 <- 1/250
r2 <- 1/5
```




```{r}
m2 <- 1.5
h <- r2*m2
m1 <- h/r1
Xo = list(b = b, c1 = c1, c2 = c2, r1 = r1, r2 = r2, m1=m1, m2=m2)
```

```{r}
eir <- rep(h/b, 3) 
```

```{r}
params = make_Xpar("hMoI", params, 1, Xo)
params = make_Xinits(params, H, 1, Xo) 
```


```{r}
fF_eir = function(eir){
  EIR = as.vector(eir) 
  return(function(t, bday=0, scale=1){scale*EIR})
} 
F_eir = fF_eir(eir)
```


```{r}
params$EIRpar = list() 
params$EIRpar$bday=0 
params$EIRpar$scale=1 
params$F_eir = F_eir
```

```{r}
params = make_indices(params)
```

```{r}
y0 <- get_inits(params)
y0
```



```{r}
out <- deSolve::ode(y = as.vector(unlist(y0)), times = c(0, 365), func = xde_derivatives, 
                    parms = params, method = 'lsoda') 
out1 <- out
```


```{r, out.width = "100%"}

colnames(out)[params$ix$X[[1]]$m1_ix+1] <- paste0('m1_', 1:params$nStrata)
colnames(out)[params$ix$X[[1]]$m2_ix+1] <- paste0('m2_', 1:params$nStrata)

out <- as.data.table(out)
out <- melt(out, id.vars = 'time')
out[, c("Component", "Strata") := tstrsplit(variable, '_', fixed = TRUE)]
out[, variable := NULL]

ggplot(data = out, mapping = aes(x = time, y = value, color = Strata)) +
  geom_line() +
  facet_wrap(Strata ~ Component, scales = 'free') +
  theme_bw()
```

# Using Setup



```{r}
xds_setup_cohort(F_eir, 'xde', Xname="hMoI", HPop=H, Xopts = Xo) ->test_hMoI_xde
```

```{r}
xde_solve(test_hMoI_xde, 365, 365)$outputs$orbits$deout -> out2
approx_equal(out2, out1)
```



