---
title: "Hybrid MoI (Multiplicity of Infection) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrid MoI (Multiplicity of Infection) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is a hybrid model which tracks the mean multiplicity of infection (superinfection) in two compartments. The first, $m1$ is all infections, and the second $m2$ are apparent (patent) infections. Therefore $m2$ is "nested" within $m1$. It is a "hybrid" model in the sense of [Nåsell (1985)](https://link.springer.com/book/10.1007/978-3-662-01609-1).

# Differential Equations

The equations are as follows:

$$
\dot{m_{1}} = h - r_{1}m_{1}
$$
$$
\dot{m_{2}} = h - r_{2}m_{2}
$$
Where $h = b EIR$, is the force of infection. Prevalence can be calculated from these MoI values by:

$$
x_{1} = 1-e^{-m_{1}}
$$
$$
x_{2} = 1-e^{-m_{2}}
$$
The net infectious probability to mosquitoes is therefore given by:

$$
x = c_{2}x_{2} + c_{1}(x_{1}-x_{2})
$$

Where $c_{1}$ is the infectiousness of inapparent infections, and $c_{2}$ is the infectiousness of patent infections.

# Equilibrium solutions

One way to proceed is assume that $m_{2}$ is known, as it models the MoI of patent (observable) infections. Then we have:

$$
h = r_{2}/m_{2}
$$
$$
m_{1} = h/r_{1}
$$
We can use this to calculate the net infectious probability, and then $\kappa = \beta^{\top} \cdot x$, allowing the equilibrium solutions of this model to feed into the other components.

# Example

```{r, message=FALSE, warning=FALSE}
library(ramp.xds)
library(deSolve)
library(viridisLite)
```

```{r, echo=FALSE} 
#devtools::load_all()
```

Here we run a simple example with 3 population strata at equilibrium. We use `ramp.xds::make_parameters_X_hMoI_xde` to
set up parameters. Please note that this only runs the human population component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

We use the null (constant) model of human demography ($H$ constant for all time). While $H$ does not appear in the equations above, it must be specified as the model of bloodfeeding ($\beta$) relies on $H$ to compute consistent values.

# The Long Way

Here, we build a model step-by-step. 

```{r}
nStrata <- 3
H <- c(100, 500, 250)
residence = rep(1,3) 
xds_obj <- make_xds_object_template("ode", "eir", 1, 1, residence)
```


```{r}
b <- 0.55
c1 <- 0.05
c2 <- 0.25
r1 <- 1/250
r2 <- 1/50
```




```{r}
m2 <- rlnorm(3, .2, .5) 
h <- r2*m2
m1 <- h/r1
Xo = list(b = b, c1 = c1, c2 = c2, r1 = r1, r2 = r2, m1=m1, m2=m2)
```

```{r}
eir <- h/b
```

```{r}
xds_obj = setup_XH_obj("hMoI", xds_obj, 1, Xo)
xds_obj = setup_XH_inits(xds_obj, H, 1, Xo) 
```

```{r}
xds_obj = setup_MY_obj("trivial", xds_obj, 1)
xds_obj = setup_L_obj("trivial", xds_obj, 1)
```


```{r}
F_season = function(t){0*t+1}
F_trend = function(t){0*t+1}
```



```{r}
xds_obj$EIRpar = list() 
xds_obj$EIRpar$eir <- as.vector(eir)
xds_obj$EIRpar$scale <- 1 
xds_obj$EIRpar$F_season <- F_season
xds_obj$EIRpar$F_trend <- F_trend
xds_obj$EIRpar$F_age <- F_flat
xds_obj$EIRpar$season_par <- list() 
xds_obj$EIRpar$trend_par<- list() 
xds_obj$EIRpar$age_par<- list() 
```

```{r}
xds_obj = make_indices(xds_obj)
```

```{r}
y0 <- get_inits(xds_obj)
y0
```



```{r}
xds_obj <- xds_solve(xds_obj) 
XH <- xds_obj$outputs$orbits$XH[[1]]
time <- xds_obj$outputs$time
```


```{r, out.width = "100%"}
clrs = turbo(5)
plot(time, XH$m1[,1], col = clrs[1], ylim = range(XH$m1), type = "l")
lines(time, XH$m1[,2], col = clrs[2])
lines(time, XH$m1[,3], col =clrs[5])
  
lines(time, XH$m2[,1], col = clrs[1], lty = 2)
lines(time, XH$m2[,2], col = clrs[2], lty=2)
lines(time, XH$m2[,3], col =clrs[5], lty=2)
```

# Using Setup



```{r}
xds_setup_eir(eir, Xname="hMoI", HPop=H, XHoptions = Xo) ->test_hMoI
```

```{r}
xds_solve(test_hMoI)-> test_hMoI 
XH2 <- test_hMoI$outputs$orbits$XH[[1]]
sum((XH$m1-XH2$m1)^2)
```



