---
title: "SIS-xde (Susceptible-Infected-Susceptible) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIS-xde (Susceptible-Infected-Susceptible) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SIS (Susceptible-Infected-Susceptible) human xde model model fulfills the generic interface of the human population component. It is the simplest model of endemic diseases in humans.

# 

We subdivide a population into susceptible ($S$) and infected and infectious ($I$) individuals, where the total population is $H = S+I.$ We assume the force of infection ($h$, FoI) is linearly proportional to the EIR: $h = b \times EIR.$ In its general form, with births ($B(H)$) and deaths (at the per-capita rate $\mu$), the generalized SIS_xde dynamics are:

$$
\begin{array}{rl}
\dot{S} &= -h S + rI + B(H) -\mu S\\
\dot{I} &= h S  - rI - \mu I
\end{array}
$$

If there is no demographic change, the SIS-xde model can be rewritten as a single equation: 

$$
\dot{I} = h (H-I) - rI
$$
Even in this simplified form, we are assuming that a population could be stratified, such that the variables and parameter are all vectors with length `nStrata.`

# Equilibrium Solutions

A typical situation when using this model is that $H$ (total population size by strata) and $X$ (number of infectious persons by strata) are known from census and survey data. Then it is of interest to find the value of $EIR$ (Entomological Inoculation Rate) which leads to that prevalence at equilibrium.

$$
0 = h \cdot (H-I) - rI 
$$

$$ 
\bar I = H \frac{h}{h+r}
$$

$$
\bar S =  H - \bar I
$$

# Example

```{r, message=FALSE, warning=FALSE}
library(ramp.xds)
library(deSolve)
library(viridisLite)
```


```{r, echo=FALSE} 
#devtools::load_all()
```

Here we run a simple example with 3 population strata at equilibrium. We use `ramp.xds::make_parameters_X_SIS_xde` to
set up parameters. Please note that this only runs the human population component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

We use the null (constant) model of human demography ($H$ constant for all time).

## The Long Way

To set up systems of differential equations, we must set the values of all our parameters. 


```{r}
nStrata <- 3
H <- c(100, 500, 250)
residence <- rep(1,3) 
nPatches=1
nHabitats=1
membership=1
params <- make_xds_template("ode", "cohort", nPatches, membership, residence)
```


```{r}
b <- rep(0.55, nStrata) 
c <- rep(0.15, nStrata) 
r <- rep(1/200, nStrata) 
Xo = list(b=b, c=c, r=r)
class(Xo) <- "SIS"
```


```{r}
foi = c(1:3)/365 
eir <- foi/b 
xde_steady_state_X(foi, H, Xo)-> ss
ss

MYZo = list(MYZm = eir*H)
```
```{r}
Xo$S=ss$S
Xo$I=ss$I
```


```{r}
params <- setup_Xpar("SIS", params, 1, Xo)
params <- setup_Xinits(params, H, 1, Xo)
params <- setup_Hpar_static(params, 1)
params <- setup_MYZpar("trivial", params, 1)
params <- setup_Lpar("trivial", params, 1)
params <- setup_Linits(params, 1)
params <- make_indices(params)
```

```{r}
F_season = function(t){0*t+1}
F_trend = function(t){0*t+1}
F_age = function(a){0*a+1}
```

```{r}
params$EIRpar = list() 
params$EIRpar$eir <- as.vector(eir)
params$EIRpar$scale <- 1 
params$EIRpar$F_season <- F_season
params$EIRpar$F_trend <- F_trend
params$EIRpar$F_age <- F_age
params <- set_eir(eir, params)
```

```{r}
params = make_indices(params)
```

```{r}
Xo$S=H 
Xo$I=H*0
params = setup_Xinits(params, H, 1, Xopts = Xo)
y0 <- get_inits(params)
y0$X
```



```{r}
params <-xds_solve_cohort(params) 
out1 <- params$outputs$orbits
```


```{r, out.width = "100%"}
clrs = turbo(5)
XH <- out1$XH[[1]] 
age <- out1$age


plot(age, XH$true_pr[,1], col = clrs[1], ylim = c(0,1), type = "l")
lines(age, XH$true_pr[,2], col = clrs[2])
lines(age, XH$true_pr[,3], col =clrs[5])
  
```


## Using Setup

We have developed utilities for setting up models. We pass the parameter values and initial values as lists: 

```{r}
xds_setup_cohort(eir, Xname="SIS", HPop=H, Xopts = Xo) -> test_SIS_xde
```

```{r}
xds_solve_cohort(test_SIS_xde)-> test_SIS_xde 
test_SIS_xde$outputs$orbits$XH[[1]] -> XH2
sum((XH$true_pr-XH2$true_pr)^2)
```

