---
title: "SEIS and SEISd" 
subtitle: "Susceptible-Exposed-Infected-Susceptible  Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEIS and SEISd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(ramp.xds)
```

```{r, echo=F}
devtools::load_all()
```

## xde 

**`ramp.xds`** includes the standard SEIS model and the delay SEIS model.   The dependent state variables for both models are:

+ $S$ the density of infected hosts

+ $E$ the density of exposed hosts who are infected but not yet infectious

+ $I$ the density of infectious hosts

In the basic versions of the model -- without demographic changes -- population density is constant, so $$H = S+E+I.$$


### Compartment Model as an ODE

The model has three parameters: 

+ $b$ is the fraction of infective bites that cause an infection;  

+ $\nu$ is the transition rate from exposed to infectious: the duration of the latent period is $1/\nu$ 

+ $r$ is the clearance rate for infections: the average duration of infection in this model is $1/r$ 

These are coupled systems of ordinary differential equations forced by the force of infection, denoted $h(t)$, where $h = F_h(E);$ here, we assume that exposure is Poisson, so we let:  

$$h = bE$$

$$
\frac{dS}{dt} = - h S + r I \\
\frac{dE}{dt} = h S - \nu E \\
\frac{dI}{dt} = \nu E - r I\\
$$

```{r}
MYZo = list(MYZm = 2000/365, f=1, q=1)
Xo = list(b= 0.5)
seis = xde_setup(Xname = "SEIS", MYZname = "trace", MYZopts = MYZo, Xopts=Xo)
seis = xde_stable_orbit(seis)
```

```{r}
unlist(list_Xvars(seis$outputs$orbits$y_last, seis, 1)) -> seis_inf
seis_inf
```

This model has the steady state... 

$$
\bar E = h S/ \nu \\ 
\bar I = \nu E / r \\ 
\bar S =  
$$
```{r}
xde_steady_state_X(1/365, 1000, seis$Xpar[[1]]) -> seis_ss
seis_ss
```

```{r}
sum((seis_inf-seis_ss)^2) < 1e-9
```

### Compartment Model as an DDE

In the delay differential equation model, we let $\nu$ denote the duration of the incubation period, and we let $h_\nu = h(t-\nu).$ 

$$
\frac{dS}{dt} = - h S + r I \\
\frac{dE}{dt} = h S - h_\nu S_\nu \\
\frac{dI}{dt} = h_\nu S_\nu - r I\\
$$
and $H = S+E+I,$  so 

$$
\bar E = h S \nu \\ 
\bar I = h S / r \\ 
\bar S =  
$$

```{r}
SEISd = xde_setup(Xname = "SEISd", MYZname = "trace", MYZopts = MYZo, Xopts=Xo)
SEISd$dlay = "dde" 
class(SEISd$dlay) = "dde" 
SEISd = xde_solve(SEISd, Tmax = 3650)
```

```{r}
unlist(list_Xvars(SEISd$outputs$orbits$y_last, SEISd, 1))[1:4] -> seisd_inf
seisd_inf
```

```{r}
xde_steady_state_X(1/365, 1000, SEISd$Xpar[[1]]) -> seisd_ss
seisd_ss
```

```{r}
sum((seisd_inf-seisd_ss)^2) < 1e-9
```

