---
title: "SIS-xde (Susceptible-Infected-Susceptible) Human Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIS-xde (Susceptible-Infected-Susceptible) Human Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The SIS (Susceptible-Infected-Susceptible) human xde model model fulfills the generic interface of the human population component. It is the simplest model of endemic diseases in humans.

# 

We subdivide a population into susceptible ($S$) and infected and infectious ($I$) individuals, where the total population is $H = S+I.$ We assume the force of infection ($h$, FoI) is linearly proportional to the EIR: $h = b \times EIR.$ In its general form, with births ($B(H)$) and deaths (at the per-capita rate $\mu$), the generalized SIS_xde dynamics are:

$$
\begin{array}{rl}
\dot{S} &= -h S + rI + B(H) -\mu S\\
\dot{I} &= h S  - rI - \mu I
\end{array}
$$

If there is no demographic change, the SIS-xde model can be rewritten as a single equation: 

$$
\dot{I} = h (H-I) - rI
$$
Even in this simplified form, we are assuming that a population could be stratified, such that the variables and parameter are all vectors with length `nStrata.`

# Equilibrium Solutions

A typical situation when using this model is that $H$ (total population size by strata) and $X$ (number of infectious persons by strata) are known from census and survey data. Then it is of interest to find the value of $EIR$ (Entomological Inoculation Rate) which leads to that prevalence at equilibrium.

$$
0 = h \cdot (H-I) - rI 
$$

$$ 
\bar I = H \frac{h}{h+r}
$$

$$
\bar S =  H - \bar I
$$

# Example

```{r, message=FALSE, warning=FALSE}
library(ramp.xds)
library(deSolve)
library(data.table)
library(ggplot2)
```


```{r, echo=FALSE} 
#devtools::load_all()
```

Here we run a simple example with 3 population strata at equilibrium. We use `ramp.xds::make_parameters_X_SIS_xde` to
set up parameters. Please note that this only runs the human population component and that most users should read [our fully worked example](ex_534.html) to run a full simulation.

We use the null (constant) model of human demography ($H$ constant for all time).

## The Long Way

To set up systems of differential equations, we must set the values of all our parameters. 


```{r}
nStrata <- 3
H <- c(100, 500, 250)
residence <- rep(1,3) 
nPatches=1
nHabitats=1
membership=1
params <- make_xds_template("xde", "cohort", "ode", nPatches, membership, residence)
```


```{r}
b <- rep(0.55, nStrata) 
c <- rep(0.15, nStrata) 
r <- rep(1/200, nStrata) 
Xo = list(b=b, c=c, r=r)
class(Xo) <- "SIS"
```


```{r}
foi = c(1:3)/365 
eir <- foi/b 
xde_steady_state_X(foi, H, Xo)-> ss
ss

MYZo = list(MYZm = eir*H)
```
```{r}
Xo$S=ss$S
Xo$I=ss$I
```



```{r}
fF_eir = function(eir){
  EIR = as.vector(eir) 
  return(function(t, bday=0, scale=1){scale*EIR})
} 
F_eir = fF_eir(eir)
```

```{r}
params$EIRpar = list() 
params$EIRpar$bday=0 
params$EIRpar$scale=1 
params$F_eir = F_eir
```

```{r}
params <- make_Xpar("SIS", params, 1, Xo)
params <- make_Xinits(params, H, 1, Xo)
params <- setup_Hpar_static(params, 1)
params <- make_MYZpar("trace", params, 1)
params <- make_Lpar("trace", params, 1)
params <- make_Linits(params, 1)
params <- make_indices(params)
```

```{r}
y0 <- as.vector(unlist(get_inits(params)))
```


```{r}
out <- deSolve::ode(y = y0, times = c(0, 365), xde_derivatives, 
                    parms = params, method = 'lsoda') 
out1 <- out
```

```{r, out.width = "100%"}
colnames(out)[params$ix$X$S_ix+1] <- paste0('S_', 1:params$nStrata)
colnames(out)[params$ix$X$I_ix+1] <- paste0('I_', 1:params$nStrata)

out <- as.data.table(out)
out <- melt(out, id.vars = 'time')
out[, c("Component", "Strata") := tstrsplit(variable, '_', fixed = TRUE)]
out[, variable := NULL]

ggplot(data = out, mapping = aes(x = time, y = value, color = Strata)) +
  geom_line() +
  facet_wrap(. ~ Component, scales = 'free') +
  theme_bw()

```

## Using Setup

We have developed utilities for setting up models. We pass the parameter values and initial values as lists: 



```{r}
xds_setup_cohort(F_eir, Xname="SIS", HPop=H, Xopts = Xo) -> test_SIS_xde
```

```{r}
xde_solve(test_SIS_xde, 365, 365)$outputs$orbits$deout -> out2
approx_equal(out2,out1)
```
