---
title: "Mosquito Spatial Ecology with `ramp.xds`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mosquito Spatial Ecology with `ramp.xds`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(ramp.xds)
```

```{r, echo=F}
#devtools::load_all()
```


In **`ramp.xds`,** mosquito spatial ecology is implemented in a patch-based simulation model: a spatial domain is subdivided into patches; and each patch contains an arbitrary number of habitats. 

Suppose there are 5 patches and 15 habitats: 5 habitats in patch 1; no habitats in patch 2; 3 habitats in patch 3; and 4 habitats in patch 4; and 2 habitats in patch 5. 

```{r}
membership = c(rep(1,5), rep(3, 3), rep(4,5), rep(5,2))
membership
```
```{r}
calJ = make_residency_matrix(5, membership)
```

The model for egg-laying is motivated by the notion of mosquito searching. Each habitat is assigned a *search weight* that is related to how easy it is for the mosquito to find it. Functions can be configured to compute mosquito *emigration* rates, $\sigma,$ as a function of habitat availability in the patch, the sum of the patch's habitat search weights.  

```{r}
set.seed(23)
searchQ = rlnorm(15, log(0.5), 1.2)
searchQ
```

```{r}
Q <- F_Q(calJ, searchQ)
```

```{r}
sigma <- exp(-Q/2)
```

```{r}
x = runif(5, -1, 1)
y = runif(5, -1, 1)
ker = function(d){exp(-d)}
K_matrix = make_K_matrix_xy(cbind(x,y), ker=ker)
K_matrix
```

```{r}
model5 <- xds_setup(nPatches=5, membership=membership, searchQ=searchQ, K_matrix=K_matrix)
```

```{r}
round(model5$ML_interface$habitat_matrix[[1]]*1000)/1000
```
