---
title: "Trace Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trace Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(ramp.xds)
library(viridisLite)
```

As part of the  *plug-and-play* modular design for **`ramp.xds`,** each dynamical component includes a trivial model that has no variables. The outputs required by other components are passed as a **trace function**. These trace functions have three parts:

+ a mean value, or a scaling argument

+ a function that returns a seasonal signal, configured with a line in the appropriate options list, `season = function(t){...}` 

+ a function that returns a trend, configured with a line in the appropriate options list, `trend = function(t){...}` 

The trivial modules each returns different values:

+ $\cal L$ - The trivial aquatic module is set up by `create_Lpar_trivial.` The function `F_emerge()` returns `Lambda*season(t)*trend(t).` To override the defaults, `Lopts` must be a named list that sets the values of the elements: 

    - `Lambda = c(...)` 
        
    - `season = function(t){...}`
        
    - `trend = function(t){...}`

+ $\cal MYZ$ - The trivial adult mosquito module is setup by `create_MYZpar_trivial.` The values will return either `F_fqZ` or `F_eggs` 

    - the function `F_fqZ()` returns `f*q*Z*season(t)*trend(t)`. To configure, `MYZopts` should be a named list that sets the values of the elements: 
    
        - `f = c(...)` 
        
        - `q = c(...)` 
        
        - `Z = c(...)`
        
        - `season = function(t){...}`
        
        - `trend = function(t){...}`
    
    - the function `F_eggs()` returns `eggs*season(t)*trend(t)`; to override the defaults, `MYZopts` must be a named list that sets the values of the elements:
    
        - ` eggs = c(...)`
        
        - `season = function(t){...}`
        
        - `trend = function(t){...}`

+ $\cal X$ - The trivial module for human / host infection and immunity is set up by `create_Xpar_trivial.` `F_X` calls `F_H` and then returns `H*kappa*season(t)*trend(t)` 

    - `F_H` is configured in `xds_setup` by passing `HPop = ...` 
    
    - To configure `F_X,` `Xopts` must be a named list that sets the values of the elements. The values of `F_X` should be in the interval $[0,H]:$ 
    
      - `kappa = c(...)`
        
      - `season = function(t){...}`
        
      - `trend = function(t){...}`
        


## Emergence 

The following adult mosquito model is forced by emergence and net infectiousness: 

+ `Lo` configures the adult mosquito emergence rate to rise initially, but then to decline with a seasonal signal: 

```{r}
Lo = list(season = function(t){1+sin(2*pi*t/365)}, 
          trend = function(t){(1+t/250)*exp(-t/1000)}, 
          Lambda =5)
```

+ `Xo` configures the net infectiousness to rise to 80% over time. 

```{r}
Xo = list(kappa = 0.8, 
          trend = function(t){exp(t/500)/(10+exp(t/500))})
```

```{r, echo=F, eval=F}
tt = seq(0, 3650, by = 10)
with(Xo, plot(tt, trend(tt), type = "l"))
```

```{r}
rm = xds_setup(Xname = "trivial", Xopts = Xo, HPop=10, 
               Lname = "trivial", Lopts = Lo)
rm <- xde_solve(rm, 3650, dt=10)
```

```{r, fig.width=6, fig.height=3.5}
xds_plot_M(rm)
xds_plot_Y(rm, add=T)
xds_plot_Z(rm, add=T)
```
