---
title: "Trace Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trace Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(ramp.xds)
library(viridisLite)
```


```{r, echo=F}
#devtools::load_all()
```

As part of the  *plug-and-play* modular design for **`ramp.xds`,** each dynamical component includes a trivial model that has no variables. The outputs required by other components are passed as a **trace function**. Trace functions in **`ramp.xds`** have three parts:

+ a mean value, and/or a scaling argument

+ a function that returns a F_seasonal signal, configured with a line in the appropriate options list, `F_season = function(t){...}` 

+ a function that returns a F_trend, configured with a line in the appropriate options list, `F_trend = function(t){...}` 

A trace function can be configured in two ways: 

+ A function can be passed as an argument;

+ Parameters to configure a function using `make_function` can be passed; by default the parameter option is set to `NULL,` and if not null, then `make_function` gets called and the parameters are saved.  



## Trivial Modules 


The trivial modules each returns different values:

+ The trivial **L**-component module is set up by `create_Lpar_trivial.` The function `F_emerge()` returns `Lambda*F_season(t)*F_trend(t).` To override the defaults, `Lopts` must be a named list that sets the values of the elements: 

    - `Lambda = c(...)` is the
        
    - `F_season = function(t){...}`
        
    - `F_trend = function(t){...}`

+ The trivial **MYZ**-component module is setup by `make_MYZpar_trivial.` The values will return either `F_fqZ` or `F_eggs` 

    - Both `F_fqZ` and `F_eggs` use the same seasonality and trend functions: 

        - `F_season = function(t){...}` 
        
        - `F_trend = function(t){...}` 

    - the function `F_fqZ()` returns `f*q*Z*F_season(t)*F_trend(t)`. To configure, `MYZopts` should be a named list that sets the values of the elements: 
    
        - `f = c(...)` 
        
        - `q = c(...)` 
        
        - `Z = c(...)`
    
    - the function `F_eggs()` returns `eggs*F_season(t)*F_trend(t)`; to override the defaults, `MYZopts` must be a named list that sets the values of the elements:
    
        - ` eggs = c(...)`

+ The trivial **X**-component module for human / host infection and immunity is set up by `create_Xpar_trivial.` `F_X` calls `F_H` and then returns `H*kappa*F_season(t)*F_trend(t)` 

    - `F_H` is configured in `xds_setup` by passing `HPop = ...` 
    
    - To configure `F_X,` `Xopts` must be a named list that sets the values of the elements. The values of `F_X` should be in the interval $[0,H]:$ 
    
      - `kappa = c(...)`
        
      - `F_season = function(t){...}`
        
      - `F_trend = function(t){...}`
        
## `make_function`

To make it easy to conduct thought experiments, **`ramp.xds`** has developed a system for setting up functions: 

  + `make_function(opts)` as a tool for generating functions with the right properties. 
  
  + `makepar_F_*` is an informal function family to construct parameter / options for `make_function` methods. 

### Seasonality

A constructor for seasonality functions drawn from a generalized family involving trigonometric functions is returned by `make_function.sin` with the associated `makepar_F_sin` that returns functions of the form: 
$$S(t) =  c\left(1 + \epsilon + \sin\left(\frac{2 \pi (t-\tau)}{365}\right)\right)^p$$

+ $c$ or `norm` is a normalizing constant 

+ $\tau$ or `phase` sets the timing of the peak

+ $\epsilon \geq 0$ or `floor` is a shape parameter: increasing the values of $\epsilon$ reduces the variance 

+ $p \geq 0$ or `pw` is a shape parameter.

```{r}
p1 = makepar_F_sin()
S1 <- make_function(p1)

```

The default normalizing constant is $365$ so that if $S$ is multiplied by some other constant, $m,$ the average daily value of the function over a year is $1.$

```{r}
integrate(S1, 0, 365)$val
```

```{r, fig.height=3.5, fig.width=5}
tt <- seq(0, 3*365, by=5)
plot(tt, S1(tt), type ="l", xlab = "Time (in Days)", ylab = expression(S(t)))
```


```{r}
p2 = makepar_F_sin(phase=120)
S2 <- make_function(p2)
```


```{r, fig.height=3.5, fig.width=5}
plot(tt, S1(tt), type ="l", xlab = "Time (in Days)", ylab = expression(S(t)))
lines(tt, S2(tt), col = "blue")
```

The function can return a vector of $N$ functions, each one configured as if $N=1$

```{r}
p3 = makepar_F_sin(phase = c(0,120), N=2)
S3 <- make_function(p3)
```

```{r, fig.height=3.5, fig.width=5}
s3 <- S3(tt) 
plot(tt, s3[1,], type ="l", xlab = "Time (in Days)", ylab = expression(S(t)))
lines(tt, S1(tt), col = "yellow", lty=2)
lines(tt, s3[2,], col = "blue")
lines(tt, S2(tt), col = "orange", lty=2)
```
```{r}
p4 <- makepar_F_sin(floor=.5)
p5 <- makepar_F_sin(floor=2)
p6 <- makepar_F_sin(pw=3)
p7 <- makepar_F_sin(pw=6)
```


```{r}
S4 <- make_function(p4)
S5 <- make_function(p5)
S6 <- make_function(p6)
S7 <- make_function(p7)
```

The shape parameters make it easy to configure a seasonality function with a range of features: 

```{r, fig.height=3.5, fig.width=5}
clrs = turbo(7)
plot(tt, S7(tt), type ="n", xlab = "Time (in Days)", ylab = expression(S(t)))
lines(tt, S1(tt), col = clrs[1])
lines(tt, S4(tt), col = clrs[2])
lines(tt, S5(tt), col = clrs[3])
lines(tt, S6(tt), col = clrs[5])
lines(tt, S7(tt), col = clrs[7])
text(1000, 3.5, expression(p==6), col=clrs[7])
text(1000, 3, expression(p==3), col=clrs[5])
text(1000, 2.5, expression(epsilon==0.5), col=clrs[2])
text(1000, 2, expression(epsilon==2), col=clrs[3])
```


###  `sigmoid` 

```{r}
ps1 <- makepar_F_sigmoid()
Fs1 <- make_function(ps1)
```

```{r, fig.height=3.5, fig.width=5}
tt <- seq(0, 365, by=5)
plot(tt, Fs1(tt), type ="l", xlab = "Time (in Days)", ylab = expression(Fs1(t)))
```

### `sharkfin`

```{r}
c1 <- makepar_F_sharkfin()
C1<- make_function(c1)
```



```{r, fig.height=3.5, fig.width=5}
tt <- seq(0, 565, by=5)
plot(tt, C1(tt), type ="l", xlab = "Time (in Days)", ylab = expression(Fs1(t)))
```



```{r}
c2a <- makepar_F_sharkfin(L = 90, dk = 1/110)
c2b <- makepar_F_sharkfin(L = 180, dk = 1/40)
```

```{r}
c2 <- makepar_F_sharkfin(L = c(90, 180), dk = c(1/110, 1/40), pw=c(2,1), N=2)
```

```{r}
C2<- make_function(c2)
```


```{r}
C2t <- C2(tt)
```



```{r, fig.height=3.5, fig.width=5}
plot(tt, C2t[1,], type ="l", xlab = "Time (in Days)", ylab = expression(Fs1(t)))
lines(tt, C2t[2,])
```

### Spline


```{r}
tv = c(-1:6)*365/4
yv = c(1/2, 1, 2, 1, 1/3, 1, 3, 1) 
sp0 <- makepar_F_splinef(tv, yv)
```


```{r}
ff <- make_function(sp0)
```


```{r}
plot(tt, ff(tt), type = "l")
```

